Reanalysis Step 1: Data Preparation

### 2025-08-19 reanalysis-1-data_preparation.qmd
wip

```{r setup}
#| include: false
# Setup analysis environment (includes seed setting)
source("dependencies.R")
setup_analysis(seed = 2025)
```

@@

```{r load-data}
# Load the combined dataset
df_combined <- readr::read_csv(reanalysis_data_file_path)
```

@@

```{r define-variables}
# Variables for propensity score model
ps_vars <- c(
  "baseline_VA_logMAR",
  "gender",
  "race_ethnicity_gp", 
  "insurance_gp",
  "age",
  "CCI",
  "DCSI",
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  "Glaucoma_bef_hitplus",
  "Otherretina_bef_hitplus",
  "cataract_surgery_bef_hitplus"
)

# Variables for outcome models
outcome_vars <- c(
  "outcome_VA_logMAR",
  "person_ever_lapse"
)

# Variables for DR severity and treatment categorization
dr_vars <- c(
  "No_DR",
  "NPDR",
  "PDR"
)

# All analysis variables
all_analysis_vars <- c(
  "e_mrn_deidentified",  # Patient ID
  ps_vars,
  outcome_vars,
  dr_vars
)
```

@@

**Modified approach for the reanalysis:**

1. **Glaucoma/Other Retina Conditions**: Missing values are recoded as 0 (absence of condition)
   - Clinical assumption: These conditions would be documented if present
   - This recoding happens BEFORE propensity score modeling
   - Maintains consistency with original analysis
   - **Is this assumption valid?** Do we want to impute these?

2. **CCI/DCSI**: **NEW - Will be imputed** (change from original analysis)
   - Original analysis excluded rows with missing CCI/DCSI
   - Reanalysis will impute these to preserve sample size and reduce bias
   - These comorbidity indices can be imputed from patient demographics and clinical characteristics

3. **Other variables**: Missing values will be handled via multiple imputation in the next step

@@

```{r missing-assessment}
# Variables used in propensity score model
ps_model_vars <- c(
  "ever_lapse_binary",
  "baseline_VA_logMAR",
  "gender_cat",
  "race_ethnic_cat", 
  "insurance_cat",
  "age_cat",
  "CCI",
  "DCSI",
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  "glaucoma_bef_hitplus_cat",  # Missing already coded as 0 (no condition)
  "otherretina_bef_hitplus_cat",  # Missing already coded as 0 (no condition)
  "catsurg_before_hitplus_cat"
)

@@

# Assess missing data BEFORE recoding (for documentation)
missing_before_recoding <- df_combined |>
  dplyr::select(dplyr::all_of(c(
    "baseline_VA_logMAR", "CCI", "DCSI", 
    "Glaucoma_bef_hitplus", "Otherretina_bef_hitplus",
    "cataract_surgery_bef_hitplus"
  ))) |>
  dplyr::summarise(dplyr::across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "n_missing_original")

@@

# Display missing data summary
missing_summary |> 
  filter(n_missing > 0) |>
  kable(caption = "Variables with missing data after recoding (glaucoma/retina conditions coded as 0)")
```

@@

```{r data-summary}
# Summary of key variables
summary_vars <- c(
  "outcome_va_vi_binary",
  "ever_lapse_binary", 
  "dr_severity",
  "treatment_type",
  "age",
  "CCI",
  "DCSI"
)

# Create summary table
df_analysis |>
  dplyr::select(dplyr::all_of(summary_vars)) |>
  summary()

# Distribution of outcome by exposure
table(df_analysis$outcome_va_vi_binary, df_analysis$ever_lapse_binary, 
      dnn = c("Vision Impairment", "Ever Lapsed"))

# Distribution of DR severity
table(df_analysis$dr_severity, useNA = "ifany")

# Distribution of treatment types
table(df_analysis$treatment_type, useNA = "ifany")
```

@@

```{r imputation-dataset}
# Full dataset for imputation (no exclusions for missing CCI/DCSI)
df_for_imputation <- df_analysis

# For comparison, create complete case dataset (as in original analysis)
df_complete_original <- df_analysis |>
  filter(!is.na(CCI) & !is.na(DCSI))

# Summary of datasets
n_total <- nrow(df_analysis)
n_complete_original <- nrow(df_complete_original)
n_gained_by_imputation <- n_total - n_complete_original

@@

```{r save-data}
# Create directories if they don't exist
if (!dir.exists(reanalysis_data_dir)) {
  dir.create(reanalysis_data_dir, recursive = TRUE)
}

# Save the prepared datasets to network drive
saveRDS(df_for_imputation, file.path(reanalysis_data_dir, "df_for_imputation.rds"))
saveRDS(df_complete_original, file.path(reanalysis_data_dir, "df_complete_original.rds"))

# Save variable lists for use in other scripts
variable_lists <- list(
  ps_model_vars = ps_model_vars,
  outcome_model_vars = outcome_model_vars,
  dr_vars = dr_vars,
  all_analysis_vars = all_analysis_vars
)
saveRDS(variable_lists, file.path(reanalysis_data_dir, "variable_lists.rds"))

# Confirm files saved
message("Data saved to: ", reanalysis_data_dir)
```

>>>>

```{r setup}
#| include: false
source("dependencies.R")
setup_analysis(seed = 2025)
```

@@

```{r load-data}
df_combined <- readr::read_csv(reanalysis_data_file_path)
```

@@

```{r define-variables}
ps_vars <- c(
  "baseline_VA_logMAR",
  "gender",
  "race_ethnicity_gp", 
  "insurance_gp",
  "age",
  "CCI",
  "DCSI",
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  "Glaucoma_bef_hitplus",
  "Otherretina_bef_hitplus",
  "cataract_surgery_bef_hitplus"
)

outcome_vars <- c(
  "outcome_VA_logMAR",
  "person_ever_lapse"
)

dr_vars <- c(
  "No_DR",
  "NPDR",
  "PDR"
)

all_analysis_vars <- c(
  "e_mrn_deidentified",  # Patient ID
  ps_vars,
  outcome_vars,
  dr_vars
)
```

@@

1. **Glaucoma/Other Retina Conditions**: Missing values are recoded as 0 (absence of condition)

- Clinical assumption: These conditions would be documented if present
- This recoding happens BEFORE propensity score modeling
- Maintains consistency with original analysis,

2. **CCI/DCSI**:

- Original analysis excluded rows with missing CCI/DCSI; this was only ~10 records
- Reanalysis will impute these using patient demographics and clinical characteristics

3. **New analysis variables**:

- This includes DR severity and treatment
- Missing values will be handled via multiple imputation in the next step

@@

```{r missing-assessment}
# Variables used in propensity score model
ps_model_vars <- c(
  "ever_lapse_binary",
  "baseline_VA_logMAR",
  "gender_cat",
  "race_ethnic_cat", 
  "insurance_cat",
  "age_cat",
  "CCI",
  "DCSI",
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  "glaucoma_bef_hitplus_cat",  # Missing coded as 0 (no condition)
  "otherretina_bef_hitplus_cat",  # Missing coded as 0 (no condition)
  "catsurg_before_hitplus_cat"
)

@@

# Assess missing data BEFORE recoding
missing_before_recoding <- df_combined |>
  dplyr::select(dplyr::all_of(c(
    "baseline_VA_logMAR", 
    "CCI", 
    "DCSI", 
    "Glaucoma_bef_hitplus",
    "Otherretina_bef_hitplus",
    "cataract_surgery_bef_hitplus"
  ))) |>
  dplyr::summarise(dplyr::across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "n_missing_original")

@@

missing_summary |> 
  filter(n_missing > 0) |>
  kable(caption = "Variables with missing data after recoding (glaucoma/retina conditions coded as 0)")
```

@@

```{r data-summary}
summary_vars <- c(
  "outcome_va_vi_binary",
  "ever_lapse_binary", 
  "dr_severity",
  "treatment_type",
  "age",
  "CCI",
  "DCSI"
)

df_analysis |>
  dplyr::select(dplyr::all_of(summary_vars)) |>
  summary()

table(df_analysis$outcome_va_vi_binary, df_analysis$ever_lapse_binary, 
      dnn = c("Vision Impairment", "Ever Lapsed"))

table(df_analysis$dr_severity, useNA = "ifany")

table(df_analysis$treatment_type, useNA = "ifany")
```

@@

```{r imputation-dataset}
df_for_imputation <- df_analysis

df_complete_original <- df_analysis |>
  filter(!is.na(CCI) & !is.na(DCSI))

n_total <- nrow(df_analysis)
n_complete_original <- nrow(df_complete_original)
n_gained_by_imputation <- n_total - n_complete_original

@@

```{r save-data}
if (!dir.exists(reanalysis_data_dir)) {
  dir.create(reanalysis_data_dir, recursive = TRUE)
}

saveRDS(df_for_imputation, file.path(reanalysis_data_dir, "df_for_imputation.rds"))
saveRDS(df_complete_original, file.path(reanalysis_data_dir, "df_complete_original.rds"))

variable_lists <- list(
  ps_model_vars = ps_model_vars,
  outcome_model_vars = outcome_model_vars,
  dr_vars = dr_vars,
  all_analysis_vars = all_analysis_vars
)
saveRDS(variable_lists, file.path(reanalysis_data_dir, "variable_lists.rds"))

message("Data saved to: ", reanalysis_data_dir)
```
